Preferences for the commodification of pensions in Chile: the rol of intergenerational social mobility

### 2025-12-01 paper/03-methods.qmd
Up borrador VF

where $Y_{it}$ indicates whether individual $i$ in wave $t$ supports more market-based pension access; $\alpha$ is the baseline probability for the reference categories; $\beta T_i$ captures the intergenerational mobility contrast: $T_i=1$ when the observed destination is $D_i=k$ (mobile to $k$) and $T_i=0$ when $D_i=j$ (immobile), so that $\beta$ is the ATT within origin $O=j (\widehat\beta=\widehat{ATT}{j,k\mid j})$, i.e., the percentage-point change in the probability of preferring pension commodification from reaching $k$ rather than remaining in $j$. The term $X_i'\gamma$ includes the pre-treatment covariates $(C)$ used to construct the entropy-balancing weights; $\lambda_t$ are wave fixed effects; and $\varepsilon_{it}$ is idiosyncratic error. The analytic sample is restricted to individuals sharing the same origin $(O=j)$. Therefore, estimation uses pooled WLS with stabilized IPW–ATT from entropy balancing and CR2 standard errors clustered by individual to address heteroskedasticity and within-person dependence.

>>>>

where $Y_{it}$ indicates whether individual $i$ in wave $t$ supports more market-based pension access; $\alpha$ is the baseline probability for the reference categories; $\beta T_i$ captures the intergenerational mobility contrast: $T_i=1$ when the observed destination is $D_i=k$ (mobile to $k$) and $T_i=0$ when $D_i=j$ (immobile), so that $\beta$ is the ATT within origin $O=j (\widehat\beta=\widehat{ATT}{j,k\mid j})$, i.e., the percentage-point change in the probability of preferring pension commodification from reaching $k$ rather than remaining in $j$. The term $X_i'\gamma$ includes the pre-treatment covariates $(C)$ used to construct the entropy-balancing weights; $\lambda_t$ are wave fixed effects; and $\varepsilon_{it}$ is idiosyncratic error. The analytic sample is restricted to individuals sharing the same origin $(O=j)$. Therefore, estimation uses pooled WLS regression with stabilized IPW–ATT from entropy balancing and CR2 standard errors clustered by individual to address heteroskedasticity and within-person dependence.

### 2025-12-01 paper/04-analysis.qmd
Up borrador VF

@fig-reg reports the double-robust estimates of the effect of intergenerational occupational mobility on preferences for pension commodification, obtained from weighted models that combine stabilized IPW with covariate adjustment and fixed wave effects. Results reveal a clear directional asymmetry between upward and downward mobility trajectories. Among upwardly mobile respondents, only the Middle→High trajectory exhibits a significant positive effect on support for pension commodification ($\beta$ = 0.09, 95% CI [0.02, 0.16]). Interpreted as an origin-specific ATT, this estimate compares—for individuals who actually moved from the middle to the high socioeconomic status—their observed endorsement with the counterfactual endorsement they would have expressed had they remained immobile in the middle socioeconomic status. Holding all other predictors constant, the point estimate implies a 9 percentage point higher probability of endorsing market-based access to pension benefits for Middle→High movers relative to their own immobility counterfactual, statistically significant at the 95% confidence level. In contrast, two downward trajectories: High→Middle ($\beta$ = −0.07, CI [−0.15, −0.00]) and High→Low ($\beta$ = −0.10, CI [−0.18, −0.02]), show significant negative effects, suggesting that individuals who experience downward mobility from higher-status origins express weaker preferences for market-based pension benefits. The remaining pathways (Low→Middle, Low→High, Middle→Low) display no significant differences relative to their non-mobile counterparts. These patterns indicate that upward movement within the upper segment (Middle→High) reinforces pro-commodification attitudes, whereas downward movement from privileged origins reduces them.

@@

To probe whether meritocratic beliefs channel the impact of mobility on preferences for pension commodification, I estimate interaction models between mobility and meritocratic beliefs and examine the conditional treatment effects of mobility at low and high levels of meritocracy (see [Supplementary Material](#anexo) for complete models).

@@

@fig-interact displays the origin-specific ATT estimates from the meritocracy interaction models as marginal effects of the mobility treatment at low and high meritocratic beliefs. Across trajectories, mobility effects among low-merit respondents are uniformly small, imprecise, and centered around zero for both upward and downward moves. Among high-merit respondents, only upward mobility from middle to high status shows a clear pro-commodification pattern, with a positive and statistically significant effect on MJP for pensions ($\beta = 0.22$, 95% CI [0.05, 0.38]). All other trajectories—Low→Middle, Low→High, Middle→Low, High→Middle, and High→Low—remain statistically indistinguishable from zero. Nonetheless, the point estimates broadly align with the theoretical expectations for upward moves: upwardly mobile individuals with stronger meritocratic beliefs tend to display higher support for pension commodification than their low-merit counterparts. By contrast, downward trajectories exhibit a pattern opposite to the mechanism hypothesis, with high-merit respondents generally showing more de-commodifying attitudes than those with weaker meritocratic beliefs. However, formal interaction tests do not detect robust heterogeneity in mobility effects by meritocratic beliefs (see [Supplementary Material](#anexo) for full interaction models): even for the Middle→High trajectory, the simple effect under high meritocracy does not differ significantly from its low-merit counterpart. Taken together, these results do not provide convincing support for the mechanism hypothesis: meritocratic beliefs do not operate as a consistent or general channel through which intergenerational mobility shapes preferences for the commodification of pensions.

>>>>

@fig-reg reports the double-robust estimates of the effect of intergenerational occupational mobility on preferences for pension commodification, obtained from weighted models that combine stabilized IPW with covariate adjustment and fixed wave effects (see [Supplementary Material](#anexo) for complete models). Results reveal a clear directional asymmetry between upward and downward mobility trajectories. Among upwardly mobile respondents, only the Middle→High trajectory exhibits a significant positive effect on support for pension commodification ($\beta$ = 0.09, 95% CI [0.02, 0.16]). Interpreted as an origin-specific ATT, this estimate compares—for individuals who actually moved from the middle to the high socioeconomic status—their observed endorsement with the counterfactual endorsement they would have expressed had they remained immobile in the middle socioeconomic status. Holding all other predictors constant, the point estimate implies a 9 percentage point higher probability of endorsing market-based access to pension benefits for Middle→High movers relative to their own immobility counterfactual, statistically significant at the 95% confidence level. In contrast, two downward trajectories: High→Middle ($\beta$ = −0.07, CI [−0.15, −0.00]) and High→Low ($\beta$ = −0.10, CI [−0.18, −0.02]), show significant negative effects, suggesting that individuals who experience downward mobility from higher-status origins express weaker preferences for market-based pension benefits. The remaining pathways (Low→Middle, Low→High, Middle→Low) display no significant differences relative to their non-mobile counterparts. These patterns indicate that upward movement within the upper segment (Middle→High) reinforces pro-commodification attitudes, whereas downward movement from privileged origins reduces them, providing favourable evidence for the first hypothesis ($H_1$).

@@

To probe whether meritocratic perceptions channel the impact of mobility on preferences for pension commodification, I estimate interaction models between mobility and meritocratic perceptions and examine the conditional treatment effects of mobility at low and high levels of meritocracy (see [Supplementary Material](#anexo) for complete models).

@@

@fig-interact displays the origin-specific ATT estimates from the meritocracy interaction models as marginal effects of the mobility treatment at low and high meritocratic perceptions. Across trajectories, mobility effects among low-merit respondents are uniformly small, imprecise, and centered around zero for both upward and downward moves. Among high-merit respondents, only upward mobility from middle to high status shows a clear pro-commodification pattern, with a positive and statistically significant effect on preferences for pension commodification ($\beta = 0.22$, 95% CI [0.05, 0.38]). All other trajectories—Low→Middle, Low→High, Middle→Low, High→Middle, and High→Low—remain statistically indistinguishable from zero. Nonetheless, the point estimates broadly align with the theoretical expectations for upward moves: upwardly mobile individuals with stronger meritocratic perceptions tend to display higher support for pension commodification than their low-merit counterparts. By contrast, downward trajectories exhibit a pattern opposite to the mechanism hypothesis, with high-merit respondents generally showing more de-commodifying attitudes than those with weaker meritocratic perceptions. However, formal interaction tests do not detect robust heterogeneity in mobility effects by meritocratic perceptions (see [Supplementary Material](#anexo) for full interaction models): even for the Middle→High trajectory, the simple effect under high meritocracy does not differ significantly from its low-merit counterpart. Taken together, these results provide no convincing evidence that meritocratic perceptions moderate mobility effects on preferences for pension commodification. Accordingly, hypotheses $H_{2a}$ and $H_{2b}$—which posit stronger pro-market reactions to upward mobility and attenuated de-commodifying reactions to downward mobility under high meritocracy—are not supported.

### 2025-12-01 paper/06-supplementary-material.qmd
Up borrador VF

```{r}
#| label: tbl-complete3
#| tbl-cap: "Marginal effects of mobility by meritocratic perception (Low vs High Meritocracy)."
#| tbl-cap-location: top
#| results: asis
#| echo: false

@@

```{r}
#| label: tbl-complete4
#| tbl-cap: "Marginal effects of mobility by survey wave (2016, 2018, and 2023)."
#| tbl-cap-location: top
#| results: asis
#| echo: false

df_tab <- df_pension_lab_t %>%
  mutate(
    Trajectory = sub("\\n.*", "", name),
    Wave = term,
    Estimate = round(estimate, 3),
    `Std. Error` = round(std.error, 3),
    `CI Low` = round(conf.low, 3),
    `CI High` = round(conf.high, 3),
  ) %>%
  select(Trajectory, Wave, Estimate, `Std. Error`, `CI Low`, `CI High`)

kableExtra::kable(df_tab,
      format = table_format,
      caption = NULL,
      align = "lcccccc") %>%
  kableExtra::kable_classic(full_width = F, html_font = "Cambria")

>>>>

```{r}
#| label: tbl-complete3
#| tbl-cap: "Interactions effects between intergenerational occupational mobility and perceived meritocracy on preferences for pension commodification, with covariates and wave fixed effects."
#| tbl-cap-location: top
#| results: asis
#| echo: false

ccoef <- list(
  "(Intercept)" = "Intercept",
  "t" = "Mobility treatment",
  "mHigh" = "High meritocracy perception (Ref.= Low)",
  "t:mHigh" = "Mobility treatment x High meritocracy perception (Ref.= Low)"
)

texreg(list(mp_lm_i,
               mp_lh_i,
               mp_mh_i,
               mp_ml_i,
               mp_hm_i,
               mp_hl_i),
          custom.model.names = c(
            "Low-Middle",
            "Low-High",
            "Middle-High",
            "Middle-Low",
            "High-Middle",
            "High-Low"
          ),
          caption.above = NULL,
          caption = NULL,
          custom.coef.map = ccoef,
          digits = 2,
          custom.note = "Note: Cells contain regression coefficients with confidence intervals in parentheses. %stars.",
          leading.zero = T,
          use.packages = F,
          booktabs = F,
          scalebox = 0.7,
          center = T,float.pos = "H!",
          custom.gof.rows = list("Controls"=rep("Yes",6)))

```{r}
#| label: tbl-complete4
#| tbl-cap: "Marginal effects of mobility by meritocratic perception (Low vs High Meritocracy)."
#| tbl-cap-location: top
#| results: asis
#| echo: false

@@
